PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes

### 2023-02-16 manuscript_00.Rmd
text updates for resubmission

---
title: 'PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes'
author:
  - Vinay Swamy:
      institute: bg
  - Zachary Batz:
      institute: nnrl
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encomopassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications. To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers. CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."

@@

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
library(tictoc)
#tic()()
#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
data_dir_vGiga <- '~/data/scEiaD_gigascience/'
data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
#data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
#data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
data_dir_vGiga <- '~/data/scEiaD_gigascience/'
data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
#data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
#data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'

@@

The modern era of single cell transcriptomics ha been marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment. This approach was pioneered by Macoksko et al. and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina. [@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a]. The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.

@@

The scEiaD pipeline is largely the same as in Swamy et al. [@swamyBuildingMegaSinglecell2021]. Very briefly, we collected the raw fastq files and quantified with the kallisto / bustools system [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019]. The quantified counts are merged into one matrix that is used for the downstream integration with scvitools (version 0.13.0) create a batch corrected lower dimensional space. The raw counts and scANVI batch corrected lower dimensional space is sent into Seurat for quality control, clustering, and 2D UMAP visualization. We use our previously published xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamyBuildingMegaSinglecell2021].

The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300. Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination on a per-study basis. Third, we altered the procedure for aligning gene names across species (see below for details). Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process [@xuProbabilisticHarmonizationAnnotation2021].

@@

We found that while nearly all datasets we curated could be integrated as a whole, a small number had strange behavior in the two dimensional UMAP space and/or nearly all cells from a study were machine labelled as a single cell type. These studies were hand-removed from the resource.

After evaluating the performance of the new dataset with our scPOP package we selected the following parameters for the integration with scANVI: 15 latent dimensions, 4000 highly variable genes, and 5 epochs. The UMAP was built with a min_dist of 0.1 and the clustering used 50 nearest neighbors.

@@

As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must identify shared genes. We found homologs/orthologs by pulling from the BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference. We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus musculus*. This generates a table linked by Ensembl gene IDs and gene names. We then use a full join on non-duplicated Ensembl IDs between mouse and human. As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name. This procedure gave us 17,769 shared human and mouse genes. To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes. In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum. In cases where there was no corresponding chicken or macaque gene, we filled in zeros. After these steps we had, as expected, 17,769 genes.

@@

Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv2, DropSeq, etc.). We also, where possible, annotated retina region (e.g. macula), sex, and age or developmental stage. Samples with perturbations (genetic or treatment) were not included. To import individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell. We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels. We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level). We did retain the original published (except for the broader name normalization) labels under SubCellType.

@@

We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping using the aggregateAcrossCells function from the scuttle R package [@mccarthyScaterPreprocessingQuality2017]. This created a counts matrix which has statistical properties approximating "bulk" RNA seq experiment matrices. This allowed the use of the more mature bulk RNA-seq tooling. This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments[@squairConfrontingFalseDiscoveries2021]. If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test. Any genes with a sum of 0 counts after aggregation were removed. The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + category". Contrasts were extracted either with the target category (e.g. "Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function [@loveModeratedEstimationFold2014a]. We tested each species (Human, Macaque, Mouse) separately.

@@

## New studies and improvements to the scEiaD database

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021]. We updated the scEiaD database for the PLAE v0.91 web app in six important ways. First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021]. Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020]. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020]. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20]. As before, the pipeline's scVI's integration parameters were chosen rigorously by testing inegration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021]. Each study in the resource was hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space and a small number of outlier studies were removed from the resource.

@@

Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources', display = 'cite')`). The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data. The PanglaoDB and UCSC are resources which have ingressed a large number of published single cell datasets The UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about forty-seven thousand. However, these resources provide no curation of the metadata and process the data for each study individually. Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical. The Tabula Sapiens project is an effort to sequence all cell type transcriptomes in humans; while it only contains around around ten thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells. A more substantial downside is that the cell type labels can be insufficient - for example rods and cones are not independently identified but rather labelled as photoreceptors. The Protein Atlas resource has recently used only one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g. one can see whether a gene of interest is enriched in Cones). This is problematic as it is unclear whether signal found is specific to the one dataset they use or is common across independent species and datasets.

@@

web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15. The Protein Atlas is asterisked as while it contains independent datasets, it only has one ocular study. ')

@@

feature_table <- rbind(
  c('PLAE','Yes','Yes', 'Yes', '1136 ', 'https://plae.nei.nih.gov'),
  c('Spectacle', 'Yes' ,'No', 'No', '1500', 'https://singlecell-eye.org'),
  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
  c('PanglaoDB', 'Yes', 'No', 'Yes', '47', 'https://panglaodb.se'),
  c('Protein Atlas', 'Yes*','No', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
  as_tibble()
colnames(feature_table) <- c('Resource Name', 'Independent\nDatasets', 'Harmonization', 'Ocular and\nnon-Ocular\n Data', 'Ocular Cell\nCount\n(thousands)', 'URL')
feature_table %>% flextable() %>%  fontsize(size = 10, part = 'all') %>% autofit() %>% fit_to_width(7)#width(., width = c(1,0.75,0.75,0.75,0.75,3))
# as of 2021 10 25
#toc()()

@@

While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently[@voigtSpectacleInteractiveResource2020]. PLAE is built around the scEiaD database, which use a multi-stage pipeline to build and identify a high performing scVI-based single cell RNA eye model that is used to integrate all datasets together [@swamyBuildingMegaSinglecell2021]. This allows for gene queries to be constructed and analyzed across studies. In contrast, Spectacle, as of August 2022, is a compilation of study-level datasets and which makes queries across studies impossible. Spectacle does not contain any non-ocular datasets and the cell type labels have not been curated or harmonized between studies (for example different studies can use MG, Mueller, Müller Glia, or other variations to refer to the same cell type). Finally, in contrast to PLAE, Spectacle does not make their datasets available for download in any form.

@@

There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage. The major downside is that this approach requires a large amount of disk storage, which can make local (on a personal computer) usage difficult; the scEiaD v2022-03-22 sqlite database is about 368 GB. Second, the visualization code in PLAE (e.g. the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time. In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26). Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r supFig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`d). The speed of the PLAE app is also comparable to the CELLxGENE browser serving the Tabula Sapiens resource while providing additional functionality.

@@

Genes with few citations can be considered as novel markers. We see, for example, *Onecut3* appears to well separate the horizontal cells and has few PubMed citations. To assess whether there are any other good candidates with few citations, we produced another heatmap (`r supFig_cap(name = 'undersupported_genes', display = 'cite')`). Most of these genes are differentially expressed across more than one cell type. We hand selected *C1ql2* and *Cartpt* (amacrine), *Frzb* and *Slitrk2* (Muller glia), and *Onecut3* (horizontal) and plotted these five genes in the boxplot view to confirm specificity across the full scEiaD database (`r supFig_cap(name = 'lightlyCited_candidates', display = 'cite')`). All five look to be specific to a cell type, except for *Frzb* which also is expressed in the RPE.

@@

A great many groups have proposed cell atlases. These range from efforts focusing on a subset of cells from a single layer of a tissue (for example [@tranSingleCellProfilesRetinal2019a] ) to a huge consortium encompassing multiple organs [@consortiumTabulaSapiensSingle2021]. We propose a complementary approach wherein we use the research data from the community at large to create a meta-atlas. This approach enables a small group to produce an atlas which rivals the size from massive multi-group consortiums. Furthermore the inherent variability between the data from independent groups can be used as a feature to get insight into whether a gene-cell type signal is stable across an entire community's datasets. We briefly demonstrate how the high study diversity in scEiaD can be used to validate proposed RPC to neurogenic gene expression changes.

# Supplementary Figures and Tables

>>>>

---
title: 'PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes'
author:
  - Vinay Swamy:
      institute: cu
  - Zachary Batz:
      institute: nnrl
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - cu: Department of Biomedical Informatics, Columbia University
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encompassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications. To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers. CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."

@@

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
library(tictoc)
#tic()()
#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'

@@

The modern era of single cell transcriptomics has been marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment. This approach was pioneered by Macoksko et al. and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina [@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a]. The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.

@@

Our full scEiaD pipeline was published previously in Swamy et al. [@swamyBuildingMegaSinglecell2021]. It is composed of three Snakefiles which perform three core workflow tasks[@molderSustainableDataAnalysis2021]. The "SnakeQUANT" pipeline takes the raw fastq input and produces a gene by cell count matrix with kallisto and bustools [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019]. This matrix is used in the "SnakePOP" pipeline to identify the optimal parameters for the scANVI based batch correction process (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`). Outputs from this pipeline include the batch corrected lower dimensional space, the UMAP coordinates, cell cluster assignments, and the cell type predictions. The outputs from this pipeline are used in the "SnakeSCEIAD" which performs differential gene testing across multiple conditions and wraps the various data types into the scEiaD sqlite database for use in the plae app (`r supFig_cap(name = 'workflow_network', display = 'cite')`).

## Quality Control

To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets. We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts. To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016] to generate per cell doublet probability scores. To identify cells for removal we used the scores in two ways. First, if an individual cell had a Scrublet score greater than 0.8 (where 1.0 is 100% probability of doublet) and a doubletCell score greater than 1,000,000 (unitless measure where higher value is more likely to be doublet) the cell is labelled as a doublet. Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets. These cutoffs were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.

## Alterations from Swamy et al. to increase stringency of the dataset

The following alterations were made to Swamy et al.: First, the cutoff to retain a cell was raised from 200 quantified unique genes to 300. Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination on a per-study basis to produce a updated count matrix. Third, we altered the procedure for aligning gene names across species (see below for details). Fourth, we use the scANVI (scvitools based) batch correction method, which leverages information from the known cell type labels in the batch correction process [@xuProbabilisticHarmonizationAnnotation2021].

@@

## Optimal parameter choices for scANVI batch correction

We found in Swamy et al. that scVI had the best batch integration performance relative to other tools like Seurat's CCA, fastMNN, etc. For this study we chose to use a variant of scVI, scANVI, We also found that adjusting the number of latent dimensions, highly variable genes, and epochs had substantial effects on the integration performance. For this update of scEiaD we ran integration with 3000, 4000, 5000, and 6000 highly variable genes, 5 and 15 epochs, and 6, 8, 10, 15, and 20 scANVI outputted latent dimensions. After evaluating the performance of the new scEiaD dataset with our scPOP package we selected the following parameters for the integration with scANVI: 15 latent dimensions, 4000 highly variable genes, and 5 epochs [@swamyBuildingMegaSinglecell2021]. The UMAP was built with a min_dist of 0.1 and the clustering used 50 nearest neighbors.

While nearly all datasets we curated could be integrated as a whole, a small number either did not integrate with other data (clusters were formed which were entirely composed of one study) and/or nearly all cells from a study were machine labelled as a single cell type. These studies were hand-removed from the resource. The metadata for these samples is made available as supplementary file "samples_hand_removed.csv."

@@

As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must identify shared genes. We found homologs and orthologs by downloading from the BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference. We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus musculus*. This generates a table linked by Ensembl gene IDs and gene names. We then use a full join on non-duplicated Ensembl IDs between mouse and human. As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name. This procedure gave us 17,769 shared human and mouse genes. To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes. In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum. In cases where there was no corresponding chicken or macaque gene, we filled in zeros. After these steps we had, as expected, 17,769 genes.

@@

Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv2, DropSeq, etc.). We also, where possible, annotated retina region (e.g. macula), sex, and age or developmental stage. Samples with perturbations (genetic or treatment) were not included. To import individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell. We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels and the later are the original cell type labels. We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).

@@

We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping using the aggregateAcrossCells function from the scuttle R package [@mccarthyScaterPreprocessingQuality2017]. This created a counts matrix which has statistical properties approximating "bulk" RNA seq experiment matrices. This allowed the use of the more mature bulk RNA-seq tooling. This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021]. If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test. Any genes with a sum of 0 counts after aggregation were removed. The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + category". Contrasts were extracted either with the target category (e.g. "Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function [@loveModeratedEstimationFold2014a]. We tested each species (Human, Macaque, Mouse) separately.

@@

## New studies and processing improvements enhance the scEiaD database

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021]. We updated the scEiaD database for the PLAE v0.93 web app in six important ways. First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021]. Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020]. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020]. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20]. As before, the pipeline's scVI integration parameters were chosen rigorously by testing integration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021] (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`). Each study in the resource was hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space and a small number of outlier studies were removed from the resource.

Of the `r meta_filter %>% filter(Organ == 'Eye') %>% nrow()` ocular cells, only `r meta_filter %>% filter(Organ == 'Eye', !is.na(CellType)) %>% nrow()` had cell type labels we could recover from the publications or data deposits. To label most of the remaining cells we used our xgboost approach from Swamy et al. which we previousy found performed well when compared to other cell type transfer approaches. The overall accuracy is 96% and the confusion matrix shows the predicted cell type to published cell type recall is very high for most cell types (`r supFig_cap(name = 'sup_CT_confusion', display = 'cite')`) and that cell type labelling works well across the different species  (`r supTab_cap(name = 'sup_CT_table', display = 'cite')`). The cell types with lower labelled performance tended to be ciliary, limbal, and corneal cells (which were often labelled as fibroblasts). There are fewer unique datasets for these cell types and we expect that future runs with more data will improve this performance

@@

Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources', display = 'cite')`). The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are general purpose single cell repositories that contain some ocular data. The PanglaoDB and UCSC resources have ingressed a large number of published single cell datasets. Af of late 2022, the UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about forty-seven thousand. However, these resources  do not curate and normalize metadata terms (e.g. ensuring the same naming system for cell types). Furthermore, each study is computationally processed separately. In practice each study is silo-ed off from the rest, making comparisons across studies impractical. The Tabula Sapiens project is an effort to sequence all cell type transcriptomes in humans; while it only contains around around ten thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells. A downside is that the ocular cell type labels can be insufficient - for example rods and cones are together labelled as photoreceptors. The Protein Atlas resource has used only one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g. one can see whether a gene of interest is enriched in Cones). Protein Atlas's use of only one study is problematic as it is unclear whether signal found is specific to the one dataset they use or is common across independent species and datasets.

@@

web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15. The Protein Atlas is asterisked as while it contains independent datasets, it only has one ocular study. Export-able means whether a resource allows bulk downloads of the count matrices and metadata. Organisms refers to the number of ocular species used.')

@@

feature_table <- rbind(
  c('PLAE','Yes','Yes', 'Yes', '4','Yes',  '1136 ', 'https://plae.nei.nih.gov'),
  c('Spectacle', 'Yes' ,'No', 'No', '5','No',  '1500', 'https://singlecell-eye.org'),
  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '2','Yes',  '62', 'https://cells.ucsc.edu' ),
  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '1','Yes',  '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
  c('PanglaoDB', 'Yes', 'No', 'Yes',  '2','Yes', '47',  'https://panglaodb.se'),
  c('Protein Atlas', 'Yes*','No', 'No',  '1', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
  as_tibble()
colnames(feature_table) <- c('Resource Name', 'Independent\nDatasets', 'Harmonization', 'Export-able', 'Organisms', 'Ocular and\nnon-Ocular\n Data', 'Ocular Cell\nCount\n(thousands)', 'URL')
feature_table %>% flextable() %>%  fontsize(size = 10, part = 'all') %>% autofit() %>% fit_to_width(7)#width(., width = c(1,0.75,0.75,0.75,0.75,3))
# as of 2021 10 25
#toc()()

@@

While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently[@voigtSpectacleInteractiveResource2020]. PLAE is built around the scEiaD database, which use a multi-stage pipeline to build and identify a high performing scVI-based single cell RNA eye model that is used to integrate all datasets together [@swamyBuildingMegaSinglecell2021]. This allows for gene queries to be constructed and analyzed across studies. In contrast, Spectacle, as of August 2022, is a compilation of study-level datasets and which makes queries across studies impossible. Spectacle does not contain any non-ocular datasets and the cell type labels have not been curated or harmonized between studies (for example different studies can use MG, Mueller, Müller Glia, or other variations to refer to the same cell type). Spectacle does contain more cells than PLAE because it also contains organoid and perturbed studies. Finally, in contrast to PLAE, Spectacle does not make their datasets available for download in any form.

@@

There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage. The major downside is that this approach requires a large amount of disk storage, which can make local (on a personal computer) usage difficult; the scEiaD v2022-03-22 sqlite database is about 368 GB. Second, the visualization code in PLAE (e.g. the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time. In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26). Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter$Citation %>% unique() %>% length()` publications (`r supFig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`d). The speed of the PLAE app is also comparable to the CELLxGENE browser serving the Tabula Sapiens resource while providing additional functionality.

@@

Genes with few citations can be considered as novel markers. We see, for example, *Onecut3* appears to well separate the horizontal cells and has few PubMed citations. To assess whether there are any other good candidates with few citations, we produced another heatmap (`r supFig_cap(name = 'undersupported_genes', display = 'cite')`). Most of these genes are differentially expressed across more than one cell type. We hand selected *C1ql2* and *Cartpt* (amacrine), *Frzb* and *Slitrk2* (Muller glia), and *Onecut3* (horizontal) and plotted these five genes in the boxplot view to confirm specificity across the full scEiaD database (`r supFig_cap(name = 'lightlyCited_candidates', display = 'cite')`). All five look to be specific to a cell type, except for *FRZB* which also is expressed in the RPE.

@@

A great many groups have proposed cell atlases. These range from efforts focusing on a subset of cells from a single layer of a tissue [@tranSingleCellProfilesRetinal2019a] to a huge consortium encompassing multiple organ [@consortiumTabulaSapiensSingle2021]. We propose a complementary approach wherein we use the research data from the community at large to create a meta-atlas. This strategy enables a small group to produce an atlas which rivals the size from massive multi-group consortiums. Furthermore, the inherent variability between the data from independent groups can be used as a feature to get insight into whether a gene-cell type signal is stable across an entire community’s datasets. We use this feature to demonstrate how the high study diversity in scEiaD can be used to validate proposed RPC to neurogenic gene expression changes in a R analysis document. Our community powered single cell ocular meta-atlas enables the entire ocular community to use the wealth of research it has generated.

# Acknowledgments

This research was funded by the National Institutes of Health (National Eye Institute) intramural program. This work utilized the computational resources of the NIH HPC Biowulf cluster (http://hpc.nih.gov).
